Joint estimation of phylogenies and ancestral states is not consistent

### 2018-01-02 appendix.tex
:woot: :woot: :woot:

We now state the Hadamard representation of site split generating probabilities, following Section 8.6 of \cite{Semple2003-em}.
For each edge $e$ define the edge \emph{fidelity} for that edge as
$$
\theta(e) = 1-2p(e).
$$
For an even-sized subset of $Y\subseteq\mathcal{S}$, we define the path set $P(Y)$ as the set of edges in the path connecting both elements of $Y$.
For $n$ taxa, the probability of observing site split $A\in\mathcal{P}(\siteSplitSet)$ is
\begin{equation}
\label{eq:hadamard_probability}
p_A = \frac{1}{2^{n-1}} \ \sum_{Y \subseteq \mathcal{S} : |Y| \equiv 0 (\mathrm{mod} \ 2)} \ \left[(-1)^{|Y \cap A|} \ \prod_{e\in P(Y)} \ \theta(e) \right].
\end{equation}
By convention, we set $P(\emptyset)=\emptyset$ and $\prod_{e\in\emptyset} \ \theta(e) = 1$.
For notational convenience, let
$$
p_{\emptyset} := \Pr(\siteSplitRV=\emptyset \mid \tau_1,t),
$$
with similar definitions for the remaining generating probabilities.
Table~\ref{tab:sitepatprob} contains calculations of site pattern frequencies for our two topologies (Fig.~\ref{fig:farris-fels-top}).

\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|l|}
    \hline
$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,t)$&$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_2,t)$\\
    \hline
    $\emptyset$ & $p_{\emptyset}$   &$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
    $\{1\}$     & $p_{1}$   &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
    $\{2\}$     & $p_{2}$   &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
    $\{3\}$     & $p_{3}$   &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
    $\{1,2\}$   & $p_{12}$   &$1-x^2-y^2+x^2y^2$&$1+2xy-2xyw-x^2w-y^2w+x^2y^2$\\
    $\{1,3\}$   & $p_{13}$   &$1+x^2+y^2-4xyw+x^2y^2$&$1-2xy-2xyw+x^2w+y^2w+x^2y^2$\\
    $\{2,3\}$   & $p_{23}$   &$1-x^2-y^2+x^2y^2$&$1-2xy+2xyw-x^2w-y^2w+x^2y^2$\\
    $\{1,2,3\}$ & $p_{123}$   &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
    \hline
\end{tabular}
%EM When I came back to this after the break I found it confusing that we refer to these as p_A below (eg in proof of Theorem 1) but that notation doesn't appear in this table. Can we use it here?
\caption{Site pattern probabilities $p_{\siteSplit_j}$ on the Farris tree $\tau_1$ and the Felsenstein tree $\tau_2$ obtained using the Hadamard transform.
All values multiplied by $1/8$.}
\label{tab:sitepatprob}
\end{table}

@@

Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$, if there is a unique maximum of the log-likelihood we will have $x_1=x_2$ at the maximum.
An analogous statement holds for $y_1$ and $y_2$.
It remains to show that the joint inference procedure maximizing \eqref{eq:profile_likelihood} results in a unique estimate of $t$.
Without loss of generality, we focus on the single parameter $x_1$ and show we obtain a unique maximum when performing joint inference; similar arguments holding for the remaining parameters.
We decompose the likelihood into the entropy and partial likelihood terms as in \eqref{eq:log_likelihood_simplified}.
By Gibbs's inequality, the entropy term $\shannonDivergence_{\tau^*,t^*}(\tau^*,t)$ has a unique maximum over $t$---namely $t^*$.
For the general case of $\tilde{t}=\{x_1, y_2, x_1, y_2, w\}$, we can compute the partial likelihood $\tilde{\ell}_{\tau^*,t^*}(\tau^*, t)$ as in Table~\ref{tab:farris_likelihoods}.
%EM why is 1/2 \le p obvious?
%das: I think it's true, but might take some annoying exposition to show. However, p is clearly between zero and one, which still passes the second derivative test.
Clearly this partial likelihood will be of the form $g(x_1) = p\log(1+x_1) + (1-p)\log(1-x_1)$ with $0 \le p \le 1$, which, by the second derivative test, is concave for $x_1\in[0,1]$ and thus has a unique maximum.
Since both summands of the likelihood are concave on $x_1\in[0,1]$, the likelihood is as well, and there exists a single $x_1\in[0,1]$ maximizing the likelihood.
Given this and the symmetry of the likelihood, we let $x_1=x_2=x$ and $y_1=y_2=y$.
The Felsenstein tree does not admit this property of symmetry of function inputs, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.
%EM WOOT

@@

%EM Do we get $0 \le \tilde{p} \le 1/2$ by just adding together terms in Table 1?
%das: Yah, if you add them it's 1/8 (4-4x^2) for 0<x<1.

@@

Since $\tilde{p} = 1/8\cdot(4-4(x^*)^2)$, $0 \le \tilde{p} \le 1/2$ implying $\hat{x} \ge 0$.

@@

The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
The upper bound in \eqref{eq:farris-upper-bound} is for the ancestral state split $\{1\}$.
For the ancestral state split $\emptyset$ we see the upper bound will be the same except we lose a $(1+x)^2$ term, gain a $(1-x)^2$ term, lose a $(1-w)$ term, and gain a $(1+w)$ term.
In this case, all terms involving $w$ simplify to $\log(1+w)$, maximized at $\hat{w}=1$.
For the constants above, $a_{1}$ is the multiplier for the $\log(1+x)$ term and $a_{2}$ for the $\log(1-x)$ term meaning that exchanging the above quadratic $x$ terms yields new constants
%EM It would seem helful for the below check to express these a_1' etc in terms of tilde{p}.
%das: not quite sure what this comment means actually.
\begin{equation}
    \begin{aligned}
        a_{1}' &= a_{1}-2p_{13}, \\
        a_{2}' &= a_{2}+2p_{13}, \\
    \end{aligned}
    \label{eq:a_const_prime_x}
\end{equation}
for site split $\emptyset$ and
\begin{equation}
    \begin{aligned}
        a_{3}' &= a_{3}-2p_{13}, \\
        a_{4}' &= a_{4}+2p_{13}, \\
    \end{aligned}
    \label{eq:a_const_prime_y}
\end{equation}
for site split $\{1,2\}$.
%EM This check would make a lot more sense if you put the C^{2}_{\tau_1,\tau_1}(x^*, y^*)  and the other analogous definition first.
%EM It would also be nice to do one of the calculations to get us from a_1 world to a_1' world. Just staring at table S4 doesn't make this obvious.
%das: is this a better explanation or did you want it written out explicitly?
To bound the case of $\ancestralSplitPartition_j(\tau_1, t) = \emptyset$,
\begin{align*}
&    \ell_{\tau_1,t^*}(\tau_1, t) \\
&\qquad\le      p_{\emptyset}  \cdot\log(2)
+ a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'}
+ a_{2}'\cdot\log\frac{2a_{2}'}{a_{1}'+a_{2}'} \\
&\qquad\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}}
+ a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}}
+ \log(2) \\
&\qquad := C^{2}_{\tau_1,\tau_1}(x^*, y^*),
\end{align*}
and, for the case of $\ancestralSplitPartition_j(\tau_1, t) = \{1,2\}$,
\begin{align*}
&    \ell_{\tau_1,t^*}(\tau_1, t) \\
&\qquad\le      p_{\emptyset}  \cdot\log(2)
+ a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}}
+ a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}} \\
&\qquad\qquad+ a_{3}'\cdot\log\frac{2a_{3}'}{a_{3}'+a_{4}'}
+ a_{4}'\cdot\log\frac{2a_{4}'}{a_{3}'+a_{4}'}
+ \log(2) \\
&\qquad := C^{3}_{\tau_1,\tau_1}(x^*, y^*).
\end{align*}

@@

Now substitute in values for $p_{\emptyset}$ and $p_{13}$ from Table~\ref{tab:sitepatprob}.
Simplifying, the likelihood can be written
$$
\ell(w) = \alpha_1\log(\beta+\gamma w) + \alpha_2\log(\beta-\gamma w) + \log(1+w)
$$
where
$$
\alpha_1 := \alpha_1(x^*, y^*) = \frac{1}{8} \left(1+(x^*)^2+(y^*)^2+4x^*(y^*)^2+(x^*)^2(y^*)^2\right),
$$
$$
\alpha_2 := \alpha_2(x^*, y^*) = \frac{1}{8}\left(1+(x^*)^2+(y^*)^2-4x^*(y^*)^2+(x^*)^2(y^*)^2\right),
$$
$$
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
$$
$$
\gamma := \gamma(x^*, y^*) = 4x^*y^*.
$$
Since $x^*$ and $y^*$ fall between zero and one, we make use of the inequalities (excepting the cases of $x^*=y^*=0$ and $x^*=y^*=1$)
$$
\alpha_1 > \alpha_2
$$
and
$$
\beta > \gamma.
$$
% Not sure if I need to show this, but if x and y aren't 1 it's 1+x^2+y^2+x^2y^2 > 4xy => (1+x^2)(1+y^2) > 4xy and 1+x^2 > 2x, 1+y^2 > 2y qed
The derivative of the log-likelihood with respect to $w$ is
$$
\ell'(w) := \frac{d}{dw} \ell(w) = \frac{\alpha_1 \gamma}{\beta+\gamma w} - \frac{\alpha_2 \gamma}{\beta-\gamma w} + \frac{1}{1+w}.
$$
The inequality $\beta > \gamma$ implies that this function stays finite and that, when considering $\ell'(w) \lessgtr 0$, we equivalently consider $f(w) \lessgtr 0$ where $f$ is the quadratic function
\begin{align*}
f(w) &= (w)^2\cdot(-\gamma^2\alpha_1-\gamma^2\alpha_2-\gamma^2) \\
      &\qquad + w\cdot(\gamma\alpha_1\beta-\gamma^2\alpha_1-\gamma\alpha_2\beta-\gamma^2\alpha_2) \\
      &\qquad + (\gamma\alpha_1\beta-\gamma\alpha_2\beta+\beta^2).
\end{align*}
% Note from das: b^2 - 4ac is always \ge 0 by \alpha_1 > \alpha_2, so b^2 - 4ac can only be zero if b is zero and a or c is zero. c is never zero, so this means a must be zero for this to happen, which only occurs if \gamma = 0, i.e., if x or y is zero.
Excepting all the cases where $x^*=0$ or where $y^*=0$, this implies $\ell'$ has two zeros according to the quadratic formula and $\alpha_1 > \alpha_2$.
Also because $\alpha_1 > \alpha_2$ we have $\ell'(0) > 0$, and thus $\ell$ is increasing at $w=0$.
This implies that if the smaller of the zeros of $f(w)$ is greater than one, then $\hat{w} \equiv 1$.
Using the quadratic formula with
$$
a = -\gamma^2 \alpha_1 - \gamma^2 \alpha_2 - \gamma^2,
$$
$$
b = \gamma \alpha_1 \beta - \gamma^2\alpha_1 - \gamma \alpha_2 \beta - \gamma^2 \alpha_2,
$$
$$
c = \gamma \alpha_1 \beta - \gamma \alpha_2 \beta + \beta^2,
$$
the smaller zero is
$$
\hat{w} = \frac{-b - \sqrt{b^2 - 4ac}}{2a},
$$
which is a function of the generating parameters $x^*$ and $y^*$.
We see that $a \le 0$ and, by a small calculation, $2a+b \leq 0$.
With this we have,
$$
\hat{w} \ge 1 \iff
|2a + b| \leq \sqrt{b^2 - 4ac} \iff
%-b - \sqrt{b^2 - 4ac} \le 2a,
%0 \ge -b - 2a \le \sqrt{b^2 - 4ac},
%b^2 + 4ab + 4a^2 \le b^2 - 4ac,
%a^2 + ab + ac \le 0,
a + b + c \ge 0.
$$
Using
$$
\alpha_1 + \alpha_2 = \frac{1}{4}\beta
$$
and
$$
\alpha_1 - \alpha_2 = x^*(y^*)^2,
$$
and simplifying as functions of $\gamma$ and $\beta$ shows that $a + b + c \ge 0$ is equivalent to
\begin{equation}
\label{eq:restricted-bl-result}
%\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
%EM Something's not quite right here. If we take the previous comment as truth we should get a
%EM -\gamma^2(1 + \beta) term which is what I get via independent calculation.
%das: I missed a factor of 1/2 in the \alpha_1+\alpha_2 term. Oops. I think below is right given that.
-\gamma^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0.
\end{equation}
\end{proof}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{w-hat-empirical-01-bias}
\caption{
    %EM I don't think you mean eq:marginal_likelihood here.
    %das: Ha, you're right.
    Estimates for $\hat{w}-y^*$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \cite{Wales1997} optimizing \eqref{eq:profile_likelihood}.
    Plot focuses on $0.6 < x^*, y^* < 1$ where $\hat{w}$ is estimated to be a value different from 1.
    We do not compute the bias for the white region where $\hat{w}=1$ to preserve a useful scale for the rest of the plot.
}
\label{fig:bl-general-bias}
\end{figure}

>>>>

We now state the Hadamard representation of site split generating probabilities, following Section 8.6 of \cite{Semple2003-em}.
For each edge $e$ define the edge \emph{fidelity} for that edge as
$$
\theta(e) = 1-2p(e).
$$
For an even-sized subset of $Y\subseteq\mathcal{S}$, we define the path set $P(Y)$ as the set of edges in the path connecting both elements of $Y$.
For $n$ taxa, the probability of observing site split $A\in\mathcal{P}(\siteSplitSet)$ is
\begin{equation}
\label{eq:hadamard_probability}
p_A = \frac{1}{2^{n-1}} \ \sum_{Y \subseteq \mathcal{S} : |Y| \equiv 0 (\mathrm{mod} \ 2)} \ \left[(-1)^{|Y \cap A|} \ \prod_{e\in P(Y)} \ \theta(e) \right].
\end{equation}
By convention, we set $P(\emptyset)=\emptyset$ and $\prod_{e\in\emptyset} \ \theta(e) = 1$.
For notational convenience, let
$$
p_{\siteSplit_j} := \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,t),
$$
for any site split $\siteSplit_j$.
Table~\ref{tab:sitepatprob} contains calculations of site pattern frequencies for our two topologies (Fig.~\ref{fig:farris-fels-top}).

\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|l|}
    \hline
$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,t)$&$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_2,t)$\\
    \hline
    $\emptyset$ & $p_{\emptyset}$   &$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
    $\{1\}$     & $p_{1}$   &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
    $\{2\}$     & $p_{2}$   &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
    $\{3\}$     & $p_{3}$   &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
    $\{1,2\}$   & $p_{12}$   &$1-x^2-y^2+x^2y^2$&$1+2xy-2xyw-x^2w-y^2w+x^2y^2$\\
    $\{1,3\}$   & $p_{13}$   &$1+x^2+y^2-4xyw+x^2y^2$&$1-2xy-2xyw+x^2w+y^2w+x^2y^2$\\
    $\{2,3\}$   & $p_{23}$   &$1-x^2-y^2+x^2y^2$&$1-2xy+2xyw-x^2w-y^2w+x^2y^2$\\
    $\{1,2,3\}$ & $p_{123}$   &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
    \hline
\end{tabular}
\caption{Site pattern probabilities $p_{\siteSplit_j}$ on the Farris tree $\tau_1$ and the Felsenstein tree $\tau_2$ obtained using the Hadamard transform.
All values multiplied by $1/8$.}
\label{tab:sitepatprob}
\end{table}

@@

Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$, if there is a unique maximum of the log-likelihood we will have $x_1=x_2$ at the maximum.
An analogous statement holds for $y_1$ and $y_2$.
It remains to show that the joint inference procedure maximizing \eqref{eq:profile_likelihood} results in a unique estimate of $t$.
Without loss of generality, we focus on the single parameter $x_1$ and show we obtain a unique maximum when performing joint inference; similar arguments holding for the remaining parameters.
We decompose the likelihood into the entropy and partial likelihood terms as in \eqref{eq:log_likelihood_simplified}.
By Gibbs's inequality, the entropy term $\shannonDivergence_{\tau^*,t^*}(\tau^*,t)$ has a unique maximum over $t$---namely $t^*$.
For the general case of $\tilde{t}=\{x_1, y_2, x_1, y_2, w\}$, we can compute the partial likelihood $\tilde{\ell}_{\tau^*,t^*}(\tau^*, t)$ as in Table~\ref{tab:farris_likelihoods}.
Clearly this partial likelihood will be of the form $g(x_1) = p\log(1+x_1) + (1-p)\log(1-x_1)$ with $0 \le p \le 1$, which, by the second derivative test, is concave for $x_1\in[0,1]$ and thus has a unique maximum.
Since both summands of the likelihood are concave on $x_1\in[0,1]$, the likelihood is as well, and there exists a single $x_1\in[0,1]$ maximizing the likelihood.
Given this and the symmetry of the likelihood, we let $x_1=x_2=x$ and $y_1=y_2=y$.
The Felsenstein tree does not admit this property of symmetry of function inputs, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

@@

@@

Since $\tilde{p} = 1/8\cdot(4-4(x^*)^2)$, we have $0 \le \tilde{p} \le 1/2$ implying $\hat{x} \ge 0$.

@@

The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
The upper bound in \eqref{eq:farris-upper-bound} is for the ancestral state split $\{1\}$.
For the ancestral state split $\emptyset$ we see the upper bound will be the same except we lose a $(1+x)^2$ term, gain a $(1-x)^2$ term, lose a $(1-w)$ term, and gain a $(1+w)$ term.
In this case, all terms involving $w$ simplify to $\log(1+w)$, maximized at $\hat{w}=1$.
For the constants above, $a_{1}$ is the multiplier for the $\log(1+x)$ term and $a_{2}$ for the $\log(1-x)$ term meaning that exchanging the above quadratic $x$ terms yields new constants
\begin{equation}
    \begin{aligned}
        a_{1}' &= a_{1}-2p_{13}, \\
        a_{2}' &= a_{2}+2p_{13}, \\
    \end{aligned}
    \label{eq:a_const_prime_x}
\end{equation}
for site split $\emptyset$ and
\begin{equation}
    \begin{aligned}
        a_{3}' &= a_{3}-2p_{13}, \\
        a_{4}' &= a_{4}+2p_{13}, \\
    \end{aligned}
    \label{eq:a_const_prime_y}
\end{equation}
for site split $\{1,2\}$.
To bound the case of $\ancestralSplitPartition_j(\tau_1, t) = \emptyset$,
\begin{align*}
&    \ell_{\tau_1,t^*}(\tau_1, t) \\
&\qquad\le      p_{\emptyset}  \cdot\log(2)
+ a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'}
+ a_{2}'\cdot\log\frac{2a_{2}'}{a_{1}'+a_{2}'} \\
&\qquad\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}}
+ a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}}
+ \log(2) \\
&\qquad := C^{2}_{\tau_1,\tau_1}(x^*, y^*),
\end{align*}
and, for the case of $\ancestralSplitPartition_j(\tau_1, t) = \{1,2\}$,
\begin{align*}
&    \ell_{\tau_1,t^*}(\tau_1, t) \\
&\qquad\le      p_{\emptyset}  \cdot\log(2)
+ a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}}
+ a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}} \\
&\qquad\qquad+ a_{3}'\cdot\log\frac{2a_{3}'}{a_{3}'+a_{4}'}
+ a_{4}'\cdot\log\frac{2a_{4}'}{a_{3}'+a_{4}'}
+ \log(2) \\
&\qquad := C^{3}_{\tau_1,\tau_1}(x^*, y^*).
\end{align*}

@@

Now substitute in values for $p_{\emptyset}$ and $p_{13}$ from Table~\ref{tab:sitepatprob}.
Simplifying, the likelihood can be written
$$
\ell(w) = \alpha_1\log(\beta+\gamma w) + \alpha_2\log(\beta-\gamma w) + \log(1+w)
$$
where
$$
\alpha_1 := \alpha_1(x^*, y^*) = \frac{1}{8} \left(1+(x^*)^2+(y^*)^2+4x^*(y^*)^2+(x^*)^2(y^*)^2\right),
$$
$$
\alpha_2 := \alpha_2(x^*, y^*) = \frac{1}{8}\left(1+(x^*)^2+(y^*)^2-4x^*(y^*)^2+(x^*)^2(y^*)^2\right),
$$
$$
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
$$
$$
\gamma := \gamma(x^*, y^*) = 4x^*y^*.
$$
Since $x^*$ and $y^*$ fall between zero and one, we make use of the inequalities (excepting the cases of $x^*=y^*=0$ and $x^*=y^*=1$)
$$
\alpha_1 > \alpha_2
$$
and
$$
\beta > \gamma.
$$
% Not sure if I need to show this, but if x and y aren't 1 it's 1+x^2+y^2+x^2y^2 > 4xy => (1+x^2)(1+y^2) > 4xy and 1+x^2 > 2x, 1+y^2 > 2y qed
The derivative of the log-likelihood with respect to $w$ is
$$
\ell'(w) := \frac{d}{dw} \ell(w) = \frac{\alpha_1 \gamma}{\beta+\gamma w} - \frac{\alpha_2 \gamma}{\beta-\gamma w} + \frac{1}{1+w}.
$$
The inequality $\beta > \gamma$ implies that this function stays finite and that, when considering $\ell'(w) \lessgtr 0$, we equivalently consider $f(w) \lessgtr 0$ where $f$ is the quadratic function
\begin{align*}
f(w) &= (w)^2\cdot(-\gamma^2\alpha_1-\gamma^2\alpha_2-\gamma^2) \\
      &\qquad + w\cdot(\gamma\alpha_1\beta-\gamma^2\alpha_1-\gamma\alpha_2\beta-\gamma^2\alpha_2) \\
      &\qquad + (\gamma\alpha_1\beta-\gamma\alpha_2\beta+\beta^2).
\end{align*}
% Note from das: b^2 - 4ac is always \ge 0 by \alpha_1 > \alpha_2, so b^2 - 4ac can only be zero if b is zero and a or c is zero. c is never zero, so this means a must be zero for this to happen, which only occurs if \gamma = 0, i.e., if x or y is zero.
Excepting all the cases where $x^*=0$ or where $y^*=0$, this implies $\ell'$ has two zeros according to the quadratic formula and $\alpha_1 > \alpha_2$.
Also because $\alpha_1 > \alpha_2$ we have $\ell'(0) > 0$, and thus $\ell$ is increasing at $w=0$.
This implies that if the smaller of the zeros of $f(w)$ is greater than one, then $\hat{w} \equiv 1$.
Using the quadratic formula with
$$
a = -\gamma^2 \alpha_1 - \gamma^2 \alpha_2 - \gamma^2,
$$
$$
b = \gamma \alpha_1 \beta - \gamma^2\alpha_1 - \gamma \alpha_2 \beta - \gamma^2 \alpha_2,
$$
$$
c = \gamma \alpha_1 \beta - \gamma \alpha_2 \beta + \beta^2,
$$
the smaller zero is
$$
\hat{w} = \frac{-b - \sqrt{b^2 - 4ac}}{2a},
$$
which is a function of the generating parameters $x^*$ and $y^*$.
We see that $a \le 0$ and, by a small calculation, $2a+b \leq 0$.
With this we have,
$$
\hat{w} \ge 1 \iff
|2a + b| \leq \sqrt{b^2 - 4ac} \iff
%-b - \sqrt{b^2 - 4ac} \le 2a,
%0 \ge -b - 2a \le \sqrt{b^2 - 4ac},
%b^2 + 4ab + 4a^2 \le b^2 - 4ac,
%a^2 + ab + ac \le 0,
a + b + c \ge 0.
$$
Using
$$
\alpha_1 + \alpha_2 = \frac{1}{4}\beta
$$
and
$$
\alpha_1 - \alpha_2 = x^*(y^*)^2,
$$
and simplifying as functions of $\gamma$ and $\beta$ shows that $a + b + c \ge 0$ is equivalent to
\begin{equation}
\label{eq:restricted-bl-result}
%\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
-\gamma^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0.
\end{equation}
\end{proof}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{w-hat-empirical-01-bias}
\caption{
    Estimates for $\hat{w}-y^*$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \cite{Wales1997} optimizing \eqref{eq:profile_likelihood}.
    Plot focuses on $0.6 < x^*, y^* < 1$ where $\hat{w}$ is estimated to be a value different from 1.
    We do not compute the bias for the white region where $\hat{w}=1$ to preserve a useful scale for the rest of the plot.
}
\label{fig:bl-general-bias}
\end{figure}
